Special Constructions in Categories

### 2017-09-21 papers/talk-6/paper-6.tex

\usepackage{todonotes}

@@

\begin{proof}
  Let $X, Y \in \cat{A}$ and $P$ be a product of $X$ and $Y$ together with $p_1 \from P \to X, p_2 \from P \to Y$
  and $P'$ be another product of $X$ and $Y$ together with $p'_1 \from P' \to X, p'_2 \from P' \to Y$.
  We apply definition \ref{def:prod} multiple times and get the following commuting diagram:
  \[ \xymatrix{
    & P \ar@/_/[ldd]_{p_1} \ar@/^/[rdd]^{p_2} & \\
    & P \ar[ld]^{p_1} \ar@/_/@{.>}[dd]|{\fbar} \ar@{.>}[u]|{f} \ar[rd]_{p_2} & \\
    X & & Y \\
    & P' \ar[lu]_{p_1'} \ar@/_/@{.>}[uu]|{\ftilde} \ar@{.>}[d]|{f'} \ar[ru]^{p'_2} & \\
    & P' \ar@/^/[luu]^{p'_1} \ar@/_/[ruu]_{p'_2} &
  } \]
  for exactly one $f, f', \fbar, \ftilde$. With the inner triangles we get:
    \begin{alignat*}{2}
    p_1  \of \ftilde &= p'_1 &\qquad p_2  \of \ftilde &= p'_2 \\
    p'_1 \of \fbar   &= p_1  &       p'_2 \of \fbar   &= p_2
  \end{alignat*}
  By substitution we get:
  \begin{alignat*}{2}
    p_1   \of (\ftilde \of \fbar) &= p_1  &\qquad p_2  \of (\ftilde \of \fbar) &= p_2 \\
    p'_1  \of (\fbar \of \ftilde) &= p'_1 &       p'_2 \of (\fbar \of \ftilde) &= p'_2
  \end{alignat*}
  With the outer triangles we get:
  \begin{alignat*}{5}
    p_1  &\;\of&&\; f  &= p_1  &\qquad p_2  &\;\of&\; f  &= p_2 \\
    p'_1 &\;\of&&\; f' &= p'_1 &\qquad p'_2 &\;\of&\; f' &= p'_2
  \end{alignat*}
  Note that $f$ and $f'$ are \emph{unique} and the first equations hold for $f = 1_P$ and $f = \ftilde \of \fbar$, thus $\ftilde \of \fbar = 1_P$.
  The second equations holds for $f' = 1_{P'}$ and for $f' = \fbar \of \ftilde$, thus $\fbar \of \ftilde = 1_{P'}$.
  Thus $P$ and $P'$ are isomorphic. \qedhere
\end{proof}

@@

\begin{example}[Reversed Cartesian product]
  \label{ex:prod:cart2}
  For sets we can define another product, where we reverse the order of the elements, i.~e.:
  \begin{alignat*}{2}
    X \times' Y &:= \setMap{\pair{y}{x}}{x \in X, \; y \in Y} \\
    p'_1(y,x)   &:= x \\
    p'_2(y,x)   &:= y
  \end{alignat*}
  Now $X \times Y$ and $X \times' Y$ are isomorphic with the involution $i(a, b) := \pair{b}{a}$.
\end{example}

\begin{example}[Posets]
  \label{ex:prod:poset}
  Let $\pair{A}{\le}$ be a poset.
  A \demph{lower bound} for two objects $x, y \in A$ is a element $a \in A$ such that $a \le x$ and $a \le y$.
  A \demph{greatest lower bound} of $x, y \in A$ is a lower bound $z \in A$ of $x$ and $y$ such that for all lower bounds $a \in A$ of $x$ and $y$, $a \le z$.
  With other words: $z$ is a greatest lower bound of $x$ and $y$ if and only if
  $$z \le x \land z \le y \land \forall a \in A, a \le x \land a \le y \Rightarrow a \le z.$$
  When we see the poset $\pair{A}{\le}$ as a category $\catA$ and have a greatest lower bound $z$ of $x, y \in A$, then $z$ is the product of $x$ and $y$.
  To see this, we apply the definition \ref{def:prod}.
  Let $x, y \in \catA$ and $z \in \catA$ be a greatest lower bound of $x$ and $y$.
  Let $a \in \catA$ be another object.
  If we have $f_1 \from a \to x$ and $f_2 \from a \to y$ then by definition of $\catA$, $a \le x$ and $a \le y$ and thus $a$ is a lower bound of $x$ and $y$.
  Now because $z$ is a greatest lower bound, $a \le z$ holds and therefore there is a unique map $\fbar \from a \to z$.
  The diagram commutes trivially.
\end{example}

\begin{example}[Direct sum of vector spaces]
  \label{ex:prod:vspace}
  Let $K$ be a field and $V, W$ vector spaces over this field.
  Then we define\footnote{One should verify that the defined structure is indeed a vector field.}
  the \demph{direct sum} $V \oplus W := \bigl(V \times W, +, \cdot \bigr)$ as a new vector field with the following operations:
  \begin{alignat*}{4}
    \pair{v_1}{w_1} + \pair{v_2}{w_2} &:= \pair{v_1 + v_2}{w_1 + w_2}           &\quad& \forall v_1, v_2 \in V, \forall w_1, w_2 \in W \\
    \alpha \cdot \pair{v}{w}          &:= \pair{\alpha \cdot v}{\alpha \cdot w} &     & \forall \alpha \in K, \forall v \in V, \forall w \in W
  \end{alignat*}
  We can show that $V \oplus W$ is the product of $V$ and $W$ in $\Vect_K$.
  The proof follows the same structure as in Example \ref{ex:prod:cart}.
\end{example}

@@

\begin{definition}[Arbitrary product]
  \label{def:arb:prod}
  Let $\catA$ be a category, $I$ a set (i.~e. of indices) and $\family{X_i}{i \in I}$ be a family of objects of $\catA$.
  A product of $\family{X_i}{i \in I}$ consists of an object $P \in \catA$ and the projection maps $\family{P \toby{p_i} X_i}{i \in I}$, such that:
  For all objects $A \in \catA$ and for all families of maps $\family{A \toby{f_i} X_i}{i \in I}$ there exists a unique map $\fbar \from A \to P$, such that $p_i \of \fbar = f_i$ holds for every index $i \in I$.
\end{definition}

\begin{example}[Arbitrary Cartesian product]
  \label{ex:arb:prod:cart}
  Let $I$ be a set of indices and $\family{X_i}{i \in I}$ be a family of objects.
  An \demph{arbitrary Cartesian tuple} of $\family{X_i}{i \in I}$ consists of a family of elements $\pi = \family{\pi_i \in X_i}{i \in I}$.
  The \demph{arbitrary Cartesian product} $\prod_{i \in I}{X_i}$ is then set set of all such tuples.
  We define the projection functions $p_i(\pi) := \pi_i$ for $i \in I$.
  Let $A$ be a set and $\family{A \toby{f_i} X_i}{i \in I}$ be a family of maps.
  Then we define $\fbar(a) := \family{f_i(a)}{i \in I}$. $p_i \of \fbar = f_i$ holds by definition for $i \in I$.
  When we have another function $\ftilde \from A \to \bigl(\prod_{i \in I}{X_i}\bigr)$ with $p_i \of \ftilde = f_i$ for $i \in I$, then $\fbar = \ftilde$, since for every $i \in I$:
  $\fbar(a) \explainRel{def}= \family{f_i(a)}{i \in I} \explainRel{ass}= \family{p_i(\ftilde(a))}{i \in I} \explainRel{ext}= \ftilde(a)$.
\end{example}

Now we want to abstract the concept of the disjoint union of sets.
We will see that the generalized structure is the dual of the product.

@@

\begin{example}[Direct sum of vector spaces]
  \label{ex:coprod:vspace}
  In Example \ref{ex:prod:vspace} we have seen that the direct sum of two vector spaces $V$ and $W$ is the product of $V$ and $W$ over a field $K$ in the category $\Vect_K$.
  We will now see that it is \emph{also} the coproduct.
  In general products and coproducts are not the same as we have seen in Examples \ref{ex:prod:cart} and \ref{ex:coprod:dunion}.
  First we have to define the two coprojection functions $p_1 \from V \to V \oplus W, p_2 \from W \to V \oplus W$:
  \begin{alignat*}{2}
    p_1(x) &:= \pair{x}{0} \\
    p_2(y) &:= \pair{0}{y}
  \end{alignat*}
  Then let $A \in \Vect_K$ and $f_1 \from V \to A, f_2 \from W \to A$ be vector morphisms.
  We define the vector morphism $\fbar \from V \oplus W \to A$ with $\fbar(v, w) := f_1(v) + f_2(w)$.
  By construction and because $f_1$ and $f_2$ are linear maps we have:
  \begin{alignat*}{6}
    & \fbar(p_1(v)) ~&\explainRel{def}=&~ \fbar(v, 0) ~&\explainRel{def}=&~ f_1(v) + f_2(0) ~&\explainRel{lin}=&~ f_1(v) \\
    & \fbar(p_2(w)) ~&\explainRel{def}=&~ \fbar(0, w) ~&\explainRel{def}=&~ f_1(0) + f_2(w) ~&\explainRel{lin}=&~ f_2(w) .
  \end{alignat*}
  Now we have to show that $\fbar$ is the only vector morphism with this property.
  Let $\ftilde \from V \oplus W \to A$ be another vector morphism with $\ftilde(p_1(v)) = f_1(v)$ and $\ftilde(p_2(w)) = f_2(w)$.
  Then we have for every $v \in V$ and $w \in W$:
  $$\fbar(v,w) \explainRel{def}= f_1(v) + f_2(w)\ \explainRel{ass}= \ftilde(p_1(v)) + \ftilde(p_2(w)) \explainRel{lin}= \ftilde(p_1(v) + p_2(w)) \explainRel{def}= \ftilde(v, w).$$
  Thus $\fbar = \ftilde$ and $V \oplus W$ is indeed the product of $V$ and $W$ in the category $\Vect_K$.
\end{example}

@@

\begin{definition}[Arbitrary coproduct]
  \label{def:arb:coprod}
  Let $\catA$ be a category, $I$ a set (i.~e. of indices) and $\family{X_i}{i \in I}$ be a family of objects of $\catA$.
  A product of $\family{X_i}{i \in I}$ consists of an object $C \in \catA$ and the coprojection maps $\family{X_i \toby{p_i} C}{i \in I}$, such that:
  For all objects $A \in \catA$ and for all families of maps $\family{X_i \toby{f_i} A}{i \in I}$ there exists a unique map $\fbar \from P \to A$, such that $\fbar \of p_i = f_i$ holds for every index $i \in I$.
\end{definition}

In Example \ref{ex:arb:prod:cart} we have generalized the Cartesian product.
No we will generalize Example \ref{ex:coprod:dunion} and define the disjoint union of a a family of sets.

\begin{example}[Arbitrary disjoint union]
  \label{ex:arb:coprod:dunion}
  Let $\family{X_i}{i \in I}$ be a familiy of sets. We define the \demph{arbitrary disjoint union}
  $$\bigdunion_{i \in I}{X_i} := \bigcup_{i \in I} \setMap{\pair{x}{i}}{x \in X_i}.$$
  We define the coprojection functions $p_i \from X_i \to \bigl(\bigdunion_{i' \in I}{X_{i'}}\bigr)$ simply by $p_i(x) := \pair{x}{i}$ for every index $i \in I$.
  Let $A$ be a set and $\family{X_i \toby{f_i} A}{i \in I}$ be a family of maps. We define $\fbar \from \bigl(\bigdunion_{i \in I}{X_i}\bigr) \to A$ by $\fbar(x, i) := f_i(x)$.
  Now $\fbar \of p_i = f_i$ holds per definition for $i \in I$.
  Let $\ftilde \from \bigl(\bigdunion_{i \in I}{X_i}\bigr) \to A$ be another function such that $\ftilde \of p_i = f_i$ holds for $i \in I$. Then
  $\fbar(x, i) \explainRel{def}= f_i(x) \explainRel{ass}= \ftilde(p_i(a)) \explainRel{def}= \ftilde(x, i)$ for $\pair{x}{i} \in \bigl(\bigdunion_{i \in I}{X_i}\bigr)$ and thus $\fbar = \ftilde$.
  Therefore $\bigdunion_{i \in I}{X_i}$ is the arbitrary disjoint union of $\family{X_i}{i \in I}$.
\end{example}

@@

Solutions sets and minus cores (as defined bellow) share the same structure.
We will generalize that structure and it equalizer.

@@

We define coequalizer as the dual of the equalizer. We will see that an example for this structure are quotient sets that are generated by a equivalence relation.

@@

\begin{definition}[Symmetric reflexive transitive closure]
  \label{def:srtc}
  Let $R$ be a binary relation over a set $X$.
  $\hat{R} := R \cup R^{-1}$ is the \demph{symmetric closure} of $R$ where $R^{-1}$ is defined by
  $R^{-1} := \setMap{\pair{b}{a}}{\pair{a}{b} \in R}$.
  The \demph{symmetric transitive closure} of $R$ is $\sim := \bigcup_{n\in\nat} {\hat R}^n$
  where ${\hat R}^n$ is recursively defined with
  ${\hat R}^0 := \setMap{\pair{a}{a}}{a \in X}$ and
  ${\hat R}^{n+1} := {\hat R}^n \of \hat R = \setMap{\pair{a}{c}}{\pair{a}{b} \in {\hat R}^n \land \pair{b}{c} \in \hat R}$.
  It can be shown that $\sim$ is the least equivalence relation containing $R$.
\end{definition}

\begin{example}[Quotient sets]
  \label{ex:coequa:quotient}
  Let $X, Y$ be sets and $s, t \from X \to Y$ be functions.
  We have to give a set $C$ and a function $i$ such that the diagram
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  commutes.
  Therefore we define the relation
  $R := \setMap{\pair{s(a)}{t(a)}}{a \in X} \subseteq Y^2$.
  Let $\sim \; \subseteq Y^2$ be the symmetric reflexive transitive closure of $R$.
  Now we define $C := Y / \sim ~:= \setMap{\class y}{y \in Y}$ where $\class y := \setMap{y'}{y \sim y'}$ are the equivalence classes.
  For the function $i \from Y \to E$ we choose $i(y) := \class y$.
  Then the diagram
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  commutes because for $x \in X$, $s(x)$ and $t(x)$ are in $R$ and therefore also in $\sim$; because $s(x) \sim t(x)$ we have $i(s(x)) = \class{s(x)} = \class{t(x)} = i(t(x))$.
  Now let
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^f & A
  } $
  be a cofork.
  By induction we can show that $f$ respects the relation $\sim$, i.~e. $f(y_1) = f(y_2)$ whenever $y_1 \sim y_2$.

  To define the function $\fbar \from C \to A$ we need a choice function $\choice \cdot \from C \to Y$ that returns for every equivalence class of $\sim$ a member of it, i.~e.
  $\forall c \in C, \forall y \in Y, y \in c \Leftrightarrow y \sim \choice{c}$
  which is equivalent to $\forall c \in C, \class {\choice c} = c$.

@@

% XXX Insert exercise 6.7 as further example?

@@

In this section we will define our last two structures: pullbacks and their duals, pushouts.

@@

\begin{example}[Pushouts in sets]
  \label{ex:pushout:set}
  To give an example for pushouts in sets we somehow have to combine the coproduct (disjoint union; see Example \ref{ex:coprod:dunion}) and the coequalizer (quotient set; see Example \ref{ex:coequa:quotient}).
  Let
  $ \xymatrix{
    X & Z \ar[l]_s \ar[r]^t & Y
  } $
  be a functions and sets.
  We want to define a set $P$ and two functions $f_1 \from X \to A, f_2 \from Y \to A$, such that the diagram
  \[ \xymatrix{
    Z \ar[r]_s \ar[d]^t & X \ar[d]_{p_1} \\
    Y \ar[r]^{p_2} & P
  } \]
  commutes.
  We define $P := (X \dunion Y) / \sim$ where $\sim \; \subseteq (X \dunion Y)^2$ is the symmetric reflexive transitive closure (see definition \ref{def:srtc}) over the relation
  $R := \setMap{\pair{\pair{s(z)}{0}}{\pair{t(z)}{1}}}{z \in Z} \subseteq (X \dunion Y)^2$.
  When we choose the projection functions
  \begin{alignat*}{2}
    p_1(x) &:= \class{\pair{x}{0}} \\
    p_2(y) &:= \class{\pair{y}{1}}
  \end{alignat*}
  the diagram commutes by construction. Now let
  \[ \xymatrix{
    Z \ar[r]_s \ar[d]^t & X \ar[d]_{f_1} \\
    Y \ar[r]^{f_2} & A
  } \]
  be another pushout square. To define the function $\fbar \from P \to A$ we first define a helper function $\fbar' \from X \dunion Y \to A$.
  Therefore we use the choice function
  $\choice \cdot \from P \to X \dunion Y$:
  \begin{alignat*}{2}
    \fbar'(x, 0) &:= f_1(x) \\
    \fbar'(y, 1) &:= f_2(y) \\
    \fbar(p)     &:= \fbar'(\choice p)
  \end{alignat*}
  By induction we can show that $\fbar'$ respects the relatation $\sim$, i.~e. $\forall a,b \in X \dunion Y, a \sim b \Rightarrow \fbar'(a) = \fbar'(b)$.

>>>>

@@

\begin{proof}
  Let $X, Y \in \cat{A}$ and $P$ be a product of $X$ and $Y$ together with $p_1 \from P \to X, p_2 \from P \to Y$
  and $P'$ be another product of $X$ and $Y$ together with $p'_1 \from P' \to X, p'_2 \from P' \to Y$.
  We apply Definition \ref{def:prod} multiple times and get the following commuting diagram:
  \[ \xymatrix{
    & P \ar@/_/[ldd]_{p_1} \ar@/^/[rdd]^{p_2} & \\
    & P \ar[ld]^{p_1} \ar@/_/@{.>}[dd]|{\fbar} \ar@{.>}[u]|{f} \ar[rd]_{p_2} & \\
    X & & Y \\
    & P' \ar[lu]_{p_1'} \ar@/_/@{.>}[uu]|{\ftilde} \ar@{.>}[d]|{f'} \ar[ru]^{p'_2} & \\
    & P' \ar@/^/[luu]^{p'_1} \ar@/_/[ruu]_{p'_2} &
  } \]
  for exactly one $f, f', \fbar, \ftilde$. With the inner triangles we get:
    \begin{alignat*}{2}
    p_1  \of \ftilde &= p'_1 &\qquad p_2  \of \ftilde &= p'_2 \\
    p'_1 \of \fbar   &= p_1  &       p'_2 \of \fbar   &= p_2
  \end{alignat*}
  By substitution we get:
  \begin{alignat*}{2}
    p_1   \of (\ftilde \of \fbar) &= p_1  &\qquad p_2  \of (\ftilde \of \fbar) &= p_2 \\
    p'_1  \of (\fbar \of \ftilde) &= p'_1 &       p'_2 \of (\fbar \of \ftilde) &= p'_2
  \end{alignat*}
  With the outer triangles we get:
  \begin{alignat*}{5}
    p_1  &\;\of&&\; f  &= p_1  &\qquad p_2  &\;\of&\; f  &= p_2 \\
    p'_1 &\;\of&&\; f' &= p'_1 &\qquad p'_2 &\;\of&\; f' &= p'_2
  \end{alignat*}
  Note that $f$ and $f'$ are \emph{unique} and the first equations hold for $f = 1_P$ and $f = \ftilde \of \fbar$, thus $\ftilde \of \fbar = 1_P$.
  The second equations holds for $f' = 1_{P'}$ and for $f' = \fbar \of \ftilde$, thus $\fbar \of \ftilde = 1_{P'}$.
  Thus $P$ and $P'$ are isomorphic. \qedhere
\end{proof}

@@

\begin{example}[Reversed Cartesian product]
  \label{ex:prod:cart2}
  For sets we can define another product, where we reverse the order of the elements, i.e.:
  \begin{alignat*}{2}
    X \times' Y &:= \setMap{\pair{y}{x}}{x \in X, \; y \in Y} \\
    p'_1(y,x)   &:= x \\
    p'_2(y,x)   &:= y
  \end{alignat*}
  Now $X \times Y$ and $X \times' Y$ are obviously isomorphic.
\end{example}

\begin{example}[Posets]
  \label{ex:prod:poset}
  Let $\pair{A}{\le}$ be a poset.
  A \demph{lower bound} for two objects $x, y \in A$ is an element $a \in A$ such that $a \le x$ and $a \le y$.
  A \demph{greatest lower bound} of $x, y \in A$ is a lower bound $z \in A$ of $x$ and $y$ such that for all lower bounds $a \in A$ of $x$ and $y$, $a \le z$.
  With other words: $z$ is a greatest lower bound of $x$ and $y$ if and only if
  $$z \le x \land z \le y \land \forall a \in A, a \le x \land a \le y \Rightarrow a \le z.$$
  When we see the poset $\pair{A}{\le}$ as a category $\catA$ and have a greatest lower bound $z$ of $x, y \in A$, then $z$ is the product of $x$ and $y$.
  To see this, we apply the Definition \ref{def:prod}.
  Let $x, y \in \catA$ and $z \in \catA$ be a greatest lower bound of $x$ and $y$.
  Let $a \in \catA$ be another object.
  If we have $f_1 \from a \to x$ and $f_2 \from a \to y$ then by definition of $\catA$, $a \le x$ and $a \le y$ and thus $a$ is a lower bound of $x$ and $y$.
  Now because $z$ is a greatest lower bound, $a \le z$ holds and therefore there is a unique map $\fbar \from a \to z$.
  The diagram commutes trivially.
\end{example}

\begin{example}[Direct sum of vector spaces]
  \label{ex:prod:vspace}
  Let $K$ be a field and $V, W$ vector spaces over this field.
  Then we define\footnote{One should verify that the defined structure is indeed a vector space.}
  the \demph{direct sum} $V \oplus W := \bigl(V \times W, +, \cdot \bigr)$ as a new vector space with the following operations:
  \begin{alignat*}{4}
    \pair{v_1}{w_1} + \pair{v_2}{w_2} &:= \pair{v_1 + v_2}{w_1 + w_2}           &\quad& \forall v_1, v_2 \in V, \forall w_1, w_2 \in W \\
    \alpha \cdot \pair{v}{w}          &:= \pair{\alpha \cdot v}{\alpha \cdot w} &     & \forall \alpha \in K, \forall v \in V, \forall w \in W
  \end{alignat*}
  We can show that $V \oplus W$ is the product of $V$ and $W$ in $\Vect_K$.
  The proof follows the same structure as in Example \ref{ex:prod:cart}.
\end{example}

@@

\begin{definition}[Arbitrary product]
  \label{def:arb:prod}
  Let $\catA$ be a category, $I$ a set (i.e. of indices) and $\family{X_i}{i \in I}$ be a family of objects of $\catA$.
  A product of $\family{X_i}{i \in I}$ consists of an object $P \in \catA$ and the projection maps $\family{P \toby{p_i} X_i}{i \in I}$, such that:
  For all objects $A \in \catA$ and for all families of maps $\family{A \toby{f_i} X_i}{i \in I}$ there exists a unique map $\fbar \from A \to P$, such that $p_i \of \fbar = f_i$ holds for every index $i \in I$.
\end{definition}

\begin{example}[Arbitrary Cartesian product]
  \label{ex:arb:prod:cart}
  Let $I$ be a set of indices and $\family{X_i}{i \in I}$ be a family of objects.
  An \demph{arbitrary Cartesian tuple} of $\family{X_i}{i \in I}$ consists of a family of elements $\pi = \family{\pi_i \in X_i}{i \in I}$.
  The \demph{arbitrary Cartesian product} $\prod_{i \in I}{X_i}$ is then the set set of all such tuples.
  We define the projection functions $p_i(\pi) := \pi_i$ for $i \in I$.
  Let $A$ be a set and $\family{A \toby{f_i} X_i}{i \in I}$ be a family of maps.
  Then we define $\fbar(a) := \family{f_i(a)}{i \in I}$.
  Then $p_i \of \fbar = f_i$ holds by definition for $i \in I$.
  When we have another function $\ftilde \from A \to \bigl(\prod_{i \in I}{X_i}\bigr)$ with $p_i \of \ftilde = f_i$ for $i \in I$, then $\fbar = \ftilde$, since for every $i \in I$:
  $\fbar(a) \explainRel{def}= \family{f_i(a)}{i \in I} \explainRel{ass}= \family{p_i(\ftilde(a))}{i \in I} \explainRel{ext}= \ftilde(a)$.
\end{example}

Now we want to abstract the concept of the disjoint union of sets.
The generalized structure is the dual of the product.

@@

\begin{example}[Direct sum of vector spaces]
  \label{ex:coprod:vspace}
  In Example \ref{ex:prod:vspace} we have seen that the direct sum of two vector spaces $V$ and $W$ over a field $K$ is the product of $V$ and $W$ in the category $\Vect_K$.
  In this example we show that it is \emph{also} the coproduct.
  In general products and coproducts are not the same as we have seen in Examples \ref{ex:prod:cart} and \ref{ex:coprod:dunion}.
  First we have to define the two coprojection functions $p_1 \from V \to V \oplus W, p_2 \from W \to V \oplus W$:
  \begin{alignat*}{2}
    p_1(x) &:= \pair{x}{0} \\
    p_2(y) &:= \pair{0}{y}
  \end{alignat*}
  Then let $A \in \Vect_K$ and $f_1 \from V \to A, f_2 \from W \to A$ be vector morphisms.
  We define the vector morphism $\fbar \from V \oplus W \to A$ with $\fbar(v, w) := f_1(v) + f_2(w)$.
  By construction and because $f_1$ and $f_2$ are linear maps we have:
  \begin{alignat*}{6}
    & \fbar(p_1(v)) ~&\explainRel{def}=&~ \fbar(v, 0) ~&\explainRel{def}=&~ f_1(v) + f_2(0) ~&\explainRel{lin}=&~ f_1(v) \\
    & \fbar(p_2(w)) ~&\explainRel{def}=&~ \fbar(0, w) ~&\explainRel{def}=&~ f_1(0) + f_2(w) ~&\explainRel{lin}=&~ f_2(w) .
  \end{alignat*}
  Now we have to show that $\fbar$ is the only vector morphism with this property.
  Let $\ftilde \from V \oplus W \to A$ be another vector morphism with $\ftilde(p_1(v)) = f_1(v)$ and $\ftilde(p_2(w)) = f_2(w)$.
  Then we have for every $v \in V$ and $w \in W$:
  $$\fbar(v,w) \explainRel{def}= f_1(v) + f_2(w)\ \explainRel{ass}= \ftilde(p_1(v)) + \ftilde(p_2(w)) \explainRel{lin}= \ftilde(p_1(v) + p_2(w)) \explainRel{def}= \ftilde(v, w).$$
  Thus $\fbar = \ftilde$ and $V \oplus W$ is indeed the product of $V$ and $W$ in the category $\Vect_K$.
\end{example}

@@

\begin{definition}[Arbitrary coproduct]
  \label{def:arb:coprod}
  Let $\catA$ be a category, $I$ a set (i.e. of indices) and $\family{X_i}{i \in I}$ be a family of objects of $\catA$.
  A product of $\family{X_i}{i \in I}$ consists of an object $C \in \catA$ and the coprojection maps $\family{X_i \toby{p_i} C}{i \in I}$, such that:
  For all objects $A \in \catA$ and for all families of maps $\family{X_i \toby{f_i} A}{i \in I}$ there exists a unique map $\fbar \from P \to A$, such that $\fbar \of p_i = f_i$ holds for every index $i \in I$.
\end{definition}

In Example \ref{ex:arb:prod:cart} we have generalized the Cartesian product.
No we generalize Example \ref{ex:coprod:dunion} and define the disjoint union of a family of sets.

\begin{example}[Arbitrary disjoint union]
  \label{ex:arb:coprod:dunion}
  Let $\family{X_i}{i \in I}$ be a family of sets. We define the \demph{arbitrary disjoint union}
  $$\bigdunion_{i \in I}{X_i} := \bigcup_{i \in I} \setMap{\pair{x}{i}}{x \in X_i}.$$
  We define the coprojection functions $p_i \from X_i \to \bigl(\bigdunion_{i' \in I}{X_{i'}}\bigr)$ simply by $p_i(x) := \pair{x}{i}$ for every index $i \in I$.
  Let $A$ be a set and $\family{X_i \toby{f_i} A}{i \in I}$ be a family of maps. We define $\fbar \from \bigl(\bigdunion_{i \in I}{X_i}\bigr) \to A$ by $\fbar(x, i) := f_i(x)$.
  Now $\fbar \of p_i = f_i$ holds per definition for $i \in I$.
  Let $\ftilde \from \bigl(\bigdunion_{i \in I}{X_i}\bigr) \to A$ be another function such that $\ftilde \of p_i = f_i$ holds for $i \in I$. Then
  $\fbar(x, i) \explainRel{def}= f_i(x) \explainRel{ass}= \ftilde(p_i(a)) \explainRel{def}= \ftilde(x, i)$ for $\pair{x}{i} \in \bigl(\bigdunion_{i \in I}{X_i}\bigr)$ and thus $\fbar = \ftilde$.
  Therefore $\bigdunion_{i \in I}{X_i}$ is the arbitrary disjoint union of $\family{X_i}{i \in I}$.
\end{example}

@@

Solution sets and minus cores (as defined bellow) share the same structure.
We generalize that structure and call it \emph{equalizer}.

@@

We define coequalizers as the dual structure of equalizers.
We show that quotient sets that are generated by a equivalence relation are an example for this structure.

@@

\begin{definition}[Symmetric reflexive transitive closure]
  \label{def:srtc}
  Let $R$ be a binary relation over a set $X$.
  $\hat{R} := R \cup R^{-1}$ is the \demph{symmetric closure} of $R$ where $R^{-1}$ is defined by
  $R^{-1} := \setMap{\pair{b}{a}}{\pair{a}{b} \in R}$.
  The \demph{symmetric reflexive transitive closure} of $R$ is $\sim ~:= \bigcup_{n\in\nat} {\hat R}^n$
  where ${\hat R}^n$ is recursively defined by
  ${\hat R}^0 := \setMap{\pair{a}{a}}{a \in X}$ and
  ${\hat R}^{n+1} := {\hat R}^n \of \hat R = \setMap{\pair{a}{c}}{\pair{a}{b} \in {\hat R}^n \land \pair{b}{c} \in \hat R}$.
  It can be shown that $\sim$ is the least equivalence relation containing $R$.
\end{definition}

\begin{example}[Quotient sets]
  \label{ex:coequa:quotient}
  Let $X, Y$ be sets and $s, t \from X \to Y$ be functions.
  We have to give a set $C$ and a function $i$ such that the diagram
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  commutes.
  Therefore we define the relation
  $R := \setMap{\pair{s(a)}{t(a)}}{a \in X} \subseteq Y^2$.
  Let $\sim \; \subseteq Y^2$ be the symmetric reflexive transitive closure of $R$.
  Now we define $C := Y / \sim ~:= \setMap{\class y}{y \in Y}$ where $\class y := \setMap{y'}{y \sim y'}$ are the equivalence classes.
  For the function $i \from Y \to E$ we choose $i(y) := \class y$.
  Then the diagram
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^i & C
  } $
  commutes because for $x \in X$, $s(x)$ and $t(x)$ are in $R$ and therefore also in $\sim$; because $s(x) \sim t(x)$ we have $i(s(x)) = \class{s(x)} = \class{t(x)} = i(t(x))$.
  Now let
  $ \xymatrix{
    X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \ar[r]^f & A
  } $
  be a cofork.
  By induction we can show that $f$ respects the relation $\sim$, i.e. $f(y_1) = f(y_2)$ whenever $y_1 \sim y_2$.

  To define the function $\fbar \from C \to A$ we need a choice function $\choice \cdot \from C \to Y$ that returns for every equivalence class of $\sim$ a member of it, i.e.
  $\forall c \in C, \forall y \in Y, y \in c \Leftrightarrow y \sim \choice{c}$
  which is equivalent to $\forall c \in C, \class {\choice c} = c$.

@@

@@

In this section we define our last two structures: pullbacks and their duals, pushouts.

@@

\begin{example}[Pushouts in sets]
  \label{ex:pushout:set}
  To give an example for pushouts in sets we somehow have to combine the coproduct (disjoint union; see Example \ref{ex:coprod:dunion}) and the coequalizer (quotient set; see Example \ref{ex:coequa:quotient}).
  Let
  $ \xymatrix{
    X & Z \ar[l]_s \ar[r]^t & Y
  } $
  be a digram in $\Set$.
  We want to define a set $P$ and two functions $f_1 \from X \to A, f_2 \from Y \to A$, such that the diagram
  \[ \xymatrix{
    Z \ar[r]_s \ar[d]^t & X \ar[d]_{p_1} \\
    Y \ar[r]^{p_2} & P
  } \]
  commutes.
  We define $P := (X \dunion Y) / \sim$ where $\sim \; \subseteq (X \dunion Y)^2$ is the symmetric reflexive transitive closure (see Definition \ref{def:srtc}) over the relation
  $R := \setMap{\pair{\pair{s(z)}{0}}{\pair{t(z)}{1}}}{z \in Z} \subseteq (X \dunion Y)^2$.
  When we choose the projection functions
  \begin{alignat*}{2}
    p_1(x) &:= \class{\pair{x}{0}} \\
    p_2(y) &:= \class{\pair{y}{1}}
  \end{alignat*}
  the diagram commutes by construction. Now let
  \[ \xymatrix{
    Z \ar[r]_s \ar[d]^t & X \ar[d]_{f_1} \\
    Y \ar[r]^{f_2} & A
  } \]
  be another pushout square. To define the function $\fbar \from P \to A$ we first define a helper function $\fbar' \from X \dunion Y \to A$.
  Therefore we use the choice function
  $\choice \cdot \from P \to X \dunion Y$:
  \begin{alignat*}{2}
    \fbar'(x, 0) &:= f_1(x) \\
    \fbar'(y, 1) &:= f_2(y) \\
    \fbar(p)     &:= \fbar'(\choice p)
  \end{alignat*}
  By induction we can show that $\fbar'$ respects the relation $\sim$, i.e. $\forall a,b \in X \dunion Y, a \sim b \Rightarrow \fbar'(a) = \fbar'(b)$.
